Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-08-20 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:20	15:33:07	(Thu20.Aug)

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. It is defined as:
%
\begin{align}
\f[\beta] = \bmat{
f_0[\beta] \\
\vdots\\
f_{M-1}[\beta]
},
\end{align}
%
where
%
\begin{align}
f_m = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. At small values the windows approach the response of a rectangular window, and at large values ($>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
\end{align}
%
where
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{md}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
}.
\end{align}
%
Here $d$ is the element spacing and $\lambda$ is the wavelength. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It is needed to ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}). Since the signal-to-noise gain is constant for a given window, this normalization also proportionally increases the  incoherent noise gain.

The steering is subject to an upper steering bound specified, which will be discussed next.

@@

We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. We numerically searched for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values. An explicit expression for was finally obtained with a third degree polynomial fit:
%
\begin{align}
\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
\end{align}
%
Any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

\begin{figure*}[tbhp!]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{Typical frequency responses for windows computed by a robust MVDR. The image is of the oiltanker Holmengraa, data collected from the 32 element HISAS1030 sonar with , operated in ssectorscan mode with an oversampling factor of ~10. The window responses was computed from 40x40 pixel groups in the shadow, highlight and speckle region. The dashed red lines at -13dB and -43dB marks the peak sidelobe level of a rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight. The phase varies mostly in highlight. Full opening angle. Parameters. }\label{mvdr_selected_windows}
\end{figure*}

>>>>

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. It is defined as:
%
\begin{align*}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}
\end{align*}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. At small values the windows approach the response of a rectangular window, and at large values ($>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
\end{align}
%
where
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{md}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
}.
\end{align}
%
Here $d$ is the element spacing and $\lambda$ is the wavelength. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. We performed a numeric search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values. A third degree polynomial fit to these points is:
%
\begin{align}
\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
\end{align}
%
Any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

\begin{figure*}[tbhp!]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{Typical frequency responses for windows computed by a robust MVDR.\newline
Left:\hfill
\parbox[t]{.95\linewidth}{A sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV. Image reconstructed with MVDR with $K=1$ temporal oversampling, $L=16$ subarray size and $d=1\%$ This is a sectorscan image with full opening angle 8 times lateral oversampling.}\newline
Right:\hfill
\parbox[t]{.95\linewidth}{The window responses was computed from 40x40 pixel groups in the shadow, highlight and speckle region. The dashed red lines at -13dB and -43dB marks the peak sidelobe level of a rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight. The phase varies mostly in highlight. Full opening angle. Parameters.} }\label{mvdr_selected_windows}
\end{figure*}
